
Differential expression for heart CAGE
```{r}
options(java.parameters = c("-XX:+UseConcMarkSweepGC", "-Xmx8192m"))
library(xlsx)

library(edgeR)
library(scales)
library(factoextra)
library(RColorBrewer)

library(ggplot2)
library(ggrepel)
```


reading metadata table - Supplementary Table 1 
```{r}
metadata<-read.xlsx2("./Supplementary Table 1 - Donor demographics, sample stats.xlsx",sheetIndex =  2, header=TRUE, startRow = 2)
head(metadata)
```

reading expression table - available from GEO GSE150736
```{r}
exp_tab <- read.table("./GSE150736_rDPI_and_Enh_expression__batch_corrected__table.txt", header = T, sep='\t', quote = "", fill = T)

# all columns are ok
all(colnames(exp_tab) == metadata$id)
```

reading annotation table
```{r}
annotation<-read.xlsx2("./Supplementary Table 2 - Clusters annotation, classification.xlsx",sheetIndex =  1, header=TRUE, startRow = 57)

# keep non-overlapping TRE
annotation <- annotation[annotation$cluster_id2 %in% rownames(exp_tab), ]
all(annotation$cluster_id2 == rownames(exp_tab))
```
TRE in promoters / enhancers
```{r}
cons_class <- gsub(".*_", "", annotation$consensus_cluster_id)
promoter_tre <- annotation$cluster_id2[cons_class == "promoter"]
enhancer_tre <- annotation$cluster_id2[cons_class == "enhancer"]
```

Design matrix - healthy vs failing, atrium and ventricle
```{r}
group<-as.factor(paste(metadata$group, gsub("S", "A", substr(metadata$heart, 2,2)), sep="_"))
design <- model.matrix(~0+group)
levels(group)
```
DE for healthy/failing ventricle/atrium
```{r}
c <- DGEList(counts = exp_tab, genes = annotation[,c(8,13,10)], group = group)
c <- calcNormFactors(c, method = "TMM")
c <- estimateCommonDisp(c)
c <- estimateTagwiseDisp(c)
norm_counts.table <- t(t(c$pseudo.counts)*(c$samples$norm.factors))
```

PCA
```{r}
res.pca <-prcomp(t(log2(norm_counts.table+1)), scale = F)
p5<-fviz_pca_ind(res.pca,
                 habillage = group,
                 geom = c("point"),  
                 title = "PCA",
                 addEllipses = TRUE,
                 show.legend = T,
                 ellipse.level=0.50,
                 palette = NULL,
                 repel = T    
)
```

```{r fig.height=3,fig.width=3}
p5
```

GLM
```{r}
c <- estimateGLMTrendedDisp(c,design)
c <- estimateGLMTagwiseDisp(c,design)
fit <- glmFit(c, design)
```

DE ventricle
```{r}
m<-rep(0,ncol(design))
aa<-grep("healthy_V", colnames(design))
vv<-grep("failing_V", colnames(design))
m[aa]<- -1/length(aa) ### control
m[vv]<- 1/length(vv) ## experiment
m
lrt <- glmLRT(fit, contrast = m)
topTags(lrt, n=nrow(lrt$table))->sorted
```

Table for Volcano plot
```{r}
sorted$table->de_ventricle
de_ventricle$logP <- -log10(de_ventricle$PValue)
de_ventricle$col <- "black"
de_ventricle$col[de_ventricle$FDR < 0.05 & de_ventricle$logFC > 1] <- "#E31D1E"
de_ventricle$col[de_ventricle$FDR < 0.05 & de_ventricle$logFC < -1] <- "#35A037"
n_genes <- 8
de_ventricle$labels <- ""
de_ventricle <- de_ventricle[-grep("^MT-", de_ventricle$GeneName), ]
de_ventricle <- de_ventricle[order(de_ventricle$logFC * de_ventricle$logP, decreasing = T), ]
de_ventricle$labels[de_ventricle$GeneName!="" & !duplicated(de_ventricle$GeneName)][1:n_genes] <- de_ventricle$GeneName[de_ventricle$GeneName!="" & !duplicated(de_ventricle$GeneName)][1:n_genes]
de_ventricle <- de_ventricle[order(de_ventricle$logFC * de_ventricle$logP, decreasing = F), ]
de_ventricle$labels[de_ventricle$GeneName!="" & !duplicated(de_ventricle$GeneName)][1:n_genes] <- de_ventricle$GeneName[de_ventricle$GeneName!="" & !duplicated(de_ventricle$GeneName)][1:n_genes]
head(de_ventricle)
```
Volcano plot
```{r  fig.height=5,fig.width=5}
ma <- ggplot(data = de_ventricle, aes(y = logP , x = logFC)) +   geom_point(colour=de_ventricle$col) + theme_light()   + geom_vline(xintercept=c(-1,1), linetype="dashed", color = "gold")  +  geom_label_repel(aes(label=labels),  data=de_ventricle,max.overlaps = 5000, label.size = NA,  fill=NA)
ma
```

```{r}
pdf("./Volcano - Ventricle DE.pdf", height = 4, width = 4)
ma
dev.off()
```




Ventricle counts
```{r}
# healthy all tre / promoters / enhancers
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC > 1])
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC > 1 & de_ventricle$cluster_id2 %in% promoter_tre])
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC > 1 & de_ventricle$cluster_id2 %in% enhancer_tre])

# failing all tre / promoters / enhancers
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC < -1])
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC < -1 & de_ventricle$cluster_id2 %in% promoter_tre])
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC < -1 & de_ventricle$cluster_id2 %in% enhancer_tre])
```


DE atrium
```{r}
m<-rep(0,ncol(design))
aa<-grep("healthy_A", colnames(design))
vv<-grep("failing_A", colnames(design))
m[aa]<- -1/length(aa) ### control
m[vv]<- 1/length(vv) ## experiment
m
lrt <- glmLRT(fit, contrast = m)
topTags(lrt, n=nrow(lrt$table))->sorted
```

Table for Volcano plot, atrium
```{r}
sorted$table->de_atrium
de_atrium$logP <- -log10(de_atrium$PValue)
de_atrium$col <- "black"
de_atrium$col[de_atrium$FDR < 0.05 & de_atrium$logFC > 1] <- "#F07E1A"
de_atrium$col[de_atrium$FDR < 0.05 & de_atrium$logFC < -1] <- "#1E79B6"
n_genes <- 8
de_atrium$labels <- ""
de_atrium <- de_atrium[-grep("^MT-", de_atrium$GeneName), ]
de_atrium <- de_atrium[order(de_atrium$logFC * de_atrium$logP, decreasing = T), ]
de_atrium$labels[de_atrium$GeneName!="" & !duplicated(de_atrium$GeneName)][1:n_genes] <- de_atrium$GeneName[de_atrium$GeneName!="" & !duplicated(de_atrium$GeneName)][1:n_genes]
de_atrium <- de_atrium[order(de_atrium$logFC * de_atrium$logP, decreasing = F), ]
de_atrium$labels[de_atrium$GeneName!="" & !duplicated(de_atrium$GeneName)][1:n_genes] <- de_atrium$GeneName[de_atrium$GeneName!="" & !duplicated(de_atrium$GeneName)][1:n_genes]
head(de_atrium)
```

Volcano plot, atrium
```{r  fig.height=5,fig.width=5}
ma <- ggplot(data = de_atrium, aes(y = logP , x = logFC)) +   geom_point(colour=de_atrium$col) + theme_light()   + geom_vline(xintercept=c(-1,1), linetype="dashed", color = "gold")  +  geom_label_repel(aes(label=labels),  data=de_atrium,max.overlaps = 5000, label.size = NA,  fill=NA)
ma
```

```{r}
pdf("./Volcano - Atrium DE.pdf", height = 4, width = 4)
ma
dev.off()
```

Atrium counts
```{r}
# healthy all tre / tre in promoters / tre in enhancers
length(de_atrium$cluster_id2[de_atrium$FDR < 0.05 & de_atrium$logFC > 1])
length(de_atrium$cluster_id2[de_atrium$FDR < 0.05 & de_atrium$logFC > 1 & de_atrium$cluster_id2 %in% promoter_tre])
length(de_atrium$cluster_id2[de_atrium$FDR < 0.05 & de_atrium$logFC > 1 & de_atrium$cluster_id2 %in% enhancer_tre])

# failing all tre / tre in promoters / tre in enhancers
length(de_atrium$cluster_id2[de_atrium$FDR < 0.05 & de_atrium$logFC < -1])
length(de_atrium$cluster_id2[de_atrium$FDR < 0.05 & de_atrium$logFC < -1 & de_atrium$cluster_id2 %in% promoter_tre])
length(de_atrium$cluster_id2[de_atrium$FDR < 0.05 & de_atrium$logFC < -1 & de_atrium$cluster_id2 %in% enhancer_tre])
```



DE healthy ventricle vs atrium
```{r}
m<-rep(0,ncol(design))
aa<-grep("healthy_V", colnames(design))
vv<-grep("healthy_A", colnames(design))
m[aa]<- -1/length(aa) ### control
m[vv]<- 1/length(vv) ## experiment
m
lrt <- glmLRT(fit, contrast = m)
topTags(lrt, n=nrow(lrt$table))->sorted
```

Table for Volcano plot
```{r}
sorted$table->de_healthy
de_healthy$logP <- -log10(de_healthy$PValue)
de_healthy$col <- "black"
de_healthy$col[de_healthy$FDR < 0.05 & de_healthy$logFC > 1] <- "#1E79B6" # healthy atrium
de_healthy$col[de_healthy$FDR < 0.05 & de_healthy$logFC < -1] <- "#35A037"# healthy ventricle
n_genes <- 8
de_healthy$labels <- ""
de_healthy <- de_healthy[-grep("^MT-", de_healthy$GeneName), ]
de_healthy <- de_healthy[order(de_healthy$logFC * de_healthy$logP, decreasing = T), ]
de_healthy$labels[de_healthy$GeneName!="" & !duplicated(de_healthy$GeneName)][1:n_genes] <- de_healthy$GeneName[de_healthy$GeneName!="" & !duplicated(de_healthy$GeneName)][1:n_genes]
de_healthy <- de_healthy[order(de_healthy$logFC * de_healthy$logP, decreasing = F), ]
de_healthy$labels[de_healthy$GeneName!="" & !duplicated(de_healthy$GeneName)][1:n_genes] <- de_healthy$GeneName[de_healthy$GeneName!="" & !duplicated(de_healthy$GeneName)][1:n_genes]
head(de_healthy)
```
Volcano plot
```{r  fig.height=5,fig.width=5}
ma <- ggplot(data = de_healthy, aes(y = logP , x = logFC)) +   geom_point(colour=de_healthy$col) + theme_light()   + geom_vline(xintercept=c(-1,1), linetype="dashed", color = "gold")  +  geom_label_repel(aes(label=labels),  data=de_healthy,max.overlaps = 5000, label.size = NA,  fill=NA)
ma
```

```{r}
pdf("./Volcano - Healthy DE.pdf", height = 4, width = 4)
ma
dev.off()
```


Healthy ventricle vs atrium counts
```{r}
# Atrium  all tre / promoters / enhancers
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC > 1])
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC > 1 & de_ventricle$cluster_id2 %in% promoter_tre])
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC > 1 & de_ventricle$cluster_id2 %in% enhancer_tre])

# Ventricle all tre / promoters / enhancers
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC < -1])
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC < -1 & de_ventricle$cluster_id2 %in% promoter_tre])
length(de_ventricle$cluster_id2[de_ventricle$FDR < 0.05 & de_ventricle$logFC < -1 & de_ventricle$cluster_id2 %in% enhancer_tre])
```


design for left-right comparison
```{r}
side <- rep(NA, length(metadata$id))
side[metadata$heart %in% c("LA", "LV")] <- "left"
side[metadata$heart %in% c("RA", "RV", "RA-sSAN", "RA-iSAN")] <- "right"
table(side)
group <- as.factor(paste(side, metadata$group))
design <- model.matrix(~0+group)
levels(group)
```

DE left-right, normalization
```{r}
c <- DGEList(counts = exp_tab, genes = annotation[,c(8, 10, 13)], group = group) 
c <- calcNormFactors(c, method = "TMM")
c <- estimateGLMTrendedDisp(c,design)
c <- estimateGLMTagwiseDisp(c, design)
fit <- glmFit(c, design)
```

DE left-right
```{r}
m<-rep(0,ncol(design))
aa<-grep("left healthy", colnames(design))
vv<-grep("right healthy", colnames(design))
m[aa]<- -1/length(aa) ### control
m[vv]<- 1/length(vv) ## experiment
m
lrt <- glmLRT(fit, contrast = m)
topTags(lrt, n=nrow(lrt$table))->sorted
```

Table for Volcano plot
```{r}
sorted$table->de_side
de_side$logP <- -log10(de_side$PValue)
de_side$col <- "black"
de_side$col[de_side$FDR < 0.05 & de_side$logFC > 1] <- "#734AA0" # Right
de_side$col[de_side$FDR < 0.05 & de_side$logFC < -1] <- "#B15928"# Left
n_genes <- 8
de_side$labels <- ""
de_side <- de_side[-grep("^MT-", de_side$GeneName), ]
de_side <- de_side[order(de_side$logFC * de_side$logP, decreasing = T), ]
de_side$labels[de_side$GeneName!="" & !duplicated(de_side$GeneName)][1:n_genes] <- de_side$GeneName[de_side$GeneName!="" & !duplicated(de_side$GeneName)][1:n_genes]
de_side <- de_side[order(de_side$logFC * de_side$logP, decreasing = F), ]
de_side$labels[de_side$GeneName!="" & !duplicated(de_side$GeneName)][1:n_genes] <- de_side$GeneName[de_side$GeneName!="" & !duplicated(de_side$GeneName)][1:n_genes]
head(de_side)
```

Volcano plot
```{r  fig.height=5,fig.width=5}
ma <- ggplot(data = de_side, aes(y = logP , x = logFC)) +   geom_point(colour=de_side$col) + theme_light()   + geom_vline(xintercept=c(-1,1), linetype="dashed", color = "gold")  +  geom_label_repel(aes(label=labels),  data=de_side,max.overlaps = 5000, label.size = NA,  fill=NA)
ma
```

```{r}
pdf("./Volcano - Side DE.pdf", height = 4, width = 4)
ma
dev.off()
```


Healthy right vs left counts
```{r}
# Right  all tre / promoters / enhancers
length(de_side$cluster_id2[de_side$FDR < 0.05 & de_side$logFC > 1])
length(de_side$cluster_id2[de_side$FDR < 0.05 & de_side$logFC > 1 & de_side$cluster_id2 %in% promoter_tre])
length(de_side$cluster_id2[de_side$FDR < 0.05 & de_side$logFC > 1 & de_side$cluster_id2 %in% enhancer_tre])

# Left all tre / promoters / enhancers
length(de_side$cluster_id2[de_side$FDR < 0.05 & de_side$logFC < -1])
length(de_side$cluster_id2[de_side$FDR < 0.05 & de_side$logFC < -1 & de_side$cluster_id2 %in% promoter_tre])
length(de_side$cluster_id2[de_side$FDR < 0.05 & de_side$logFC < -1 & de_side$cluster_id2 %in% enhancer_tre])
```




design for ICM-NICM comparison
```{r}
group <- as.factor(metadata$cardiomyopathy.type)
design <- model.matrix(~0+group)
levels(group)
```

DE ICM-NICM, normalization
```{r}
c <- DGEList(counts = exp_tab, genes = annotation[,c(8, 10, 13)], group = group) 
c <- calcNormFactors(c, method = "TMM")
c <- estimateGLMTrendedDisp(c,design)
c <- estimateGLMTagwiseDisp(c, design)
fit <- glmFit(c, design)
```

DE ICM-NICM
```{r}
m<-rep(0,ncol(design))
aa<-grep("groupICM", colnames(design))
vv<-grep("groupNICM", colnames(design))
m[aa]<- -1 ### control
m[vv]<- 1 ## experiment
m
lrt <- glmLRT(fit, contrast = m)
topTags(lrt, n=nrow(lrt$table))->sorted
```

Table for Volcano plot
```{r}
sorted$table->de_cardiomyopathy
de_cardiomyopathy$logP <- -log10(de_cardiomyopathy$PValue)
de_cardiomyopathy$col <- "black"
de_cardiomyopathy$col[de_cardiomyopathy$FDR < 0.05 & de_cardiomyopathy$logFC > 1] <- brewer.pal(9, "Set3")[3] # NICM
de_cardiomyopathy$col[de_cardiomyopathy$FDR < 0.05 & de_cardiomyopathy$logFC < -1] <- brewer.pal(9, "Set3")[1]# ICM
n_genes <- 8
de_cardiomyopathy$labels <- ""
de_cardiomyopathy <- de_cardiomyopathy[-grep("^MT-", de_cardiomyopathy$GeneName), ]
de_cardiomyopathy <- de_cardiomyopathy[order(de_cardiomyopathy$logFC * de_cardiomyopathy$logP, decreasing = T), ]
de_cardiomyopathy$labels[de_cardiomyopathy$GeneName!="" & !duplicated(de_cardiomyopathy$GeneName)][1:n_genes] <- de_cardiomyopathy$GeneName[de_cardiomyopathy$GeneName!="" & !duplicated(de_cardiomyopathy$GeneName)][1:n_genes]
de_cardiomyopathy <- de_cardiomyopathy[order(de_cardiomyopathy$logFC * de_cardiomyopathy$logP, decreasing = F), ]
de_cardiomyopathy$labels[de_cardiomyopathy$GeneName!="" & !duplicated(de_cardiomyopathy$GeneName)][1:n_genes] <- de_cardiomyopathy$GeneName[de_cardiomyopathy$GeneName!="" & !duplicated(de_cardiomyopathy$GeneName)][1:n_genes]
head(de_cardiomyopathy)
```

Volcano plot
```{r  fig.height=5,fig.width=5}
ma <- ggplot(data = de_cardiomyopathy, aes(y = logP , x = logFC)) +   geom_point(colour=de_cardiomyopathy$col) + theme_light()   + geom_vline(xintercept=c(-1,1), linetype="dashed", color = "gold")  +  geom_label_repel(aes(label=labels),  data=de_cardiomyopathy,max.overlaps = 5000, label.size = NA,  fill=NA)
ma
```

```{r}
pdf("./Volcano - Cardiomyopathy DE.pdf", height = 4, width = 4)
ma
dev.off()
```


Cardiomyopathy NICM vs ICM counts
```{r}
# NICM  all tre / promoters / enhancers
length(de_cardiomyopathy$cluster_id2[de_cardiomyopathy$FDR < 0.05 & de_cardiomyopathy$logFC > 1])
length(de_cardiomyopathy$cluster_id2[de_cardiomyopathy$FDR < 0.05 & de_cardiomyopathy$logFC > 1 & de_cardiomyopathy$cluster_id2 %in% promoter_tre])
length(de_cardiomyopathy$cluster_id2[de_cardiomyopathy$FDR < 0.05 & de_cardiomyopathy$logFC > 1 & de_cardiomyopathy$cluster_id2 %in% enhancer_tre])

# ICM all tre / promoters / enhancers
length(de_cardiomyopathy$cluster_id2[de_cardiomyopathy$FDR < 0.05 & de_cardiomyopathy$logFC < -1])
length(de_cardiomyopathy$cluster_id2[de_cardiomyopathy$FDR < 0.05 & de_cardiomyopathy$logFC < -1 & de_cardiomyopathy$cluster_id2 %in% promoter_tre])
length(de_cardiomyopathy$cluster_id2[de_cardiomyopathy$FDR < 0.05 & de_cardiomyopathy$logFC < -1 & de_cardiomyopathy$cluster_id2 %in% enhancer_tre])
```


Failing Atrium groups

```{r}
bad_samples <- c("AA25514", "AA23683", "AA24806", "AA24941", "AA23819", "AA24899", "AA24870", "AA25314", "AA23996","AA24205")
good_samples <- c("AA25197", "AA24132", "AA24852", "AA25216", "AA25002")
failing_atrium_repro <- metadata$id[ gsub(".*_", "",  metadata$id) %in%  bad_samples]
failing_atrium_control <- metadata$id[ gsub(".*_", "",  metadata$id) %in%  good_samples]
failing_ventricle_samples <- metadata$id[metadata$heart %in% c("LV", "RV") & metadata$group == "failing"]
healthy_ventricle_samples <- metadata$id[metadata$heart %in% c("LV", "RV") & metadata$group == "healthy"]
healthy_atrium_samples <- metadata$id[metadata$heart %in% c("LA", "RA", "RA-sSAN", "RA-iSAN") & metadata$group == "healthy"]
metadata$group_2 <- "group_2"
metadata$group_2[metadata$id %in% failing_atrium_control] <- "failing_atrium_control"
metadata$group_2[metadata$id %in% failing_atrium_repro] <- "failing_atrium_repro"
metadata$group_2[metadata$id %in% failing_ventricle_samples] <- "failing_ventricle"
metadata$group_2[metadata$id %in% healthy_atrium_samples] <- "healthy_atrium"
metadata$group_2[metadata$id %in% healthy_ventricle_samples] <- "healthy_ventricle"
table(metadata$group_2)
```

```{r}
group <- as.factor(metadata$group_2)
design <- model.matrix(~0+group)
c <- DGEList(counts = exp_tab, genes = annotation[,c(8, 10, 13)], group = group)
c <- calcNormFactors(c, method = "TMM")
c <- estimateGLMTrendedDisp(c,design)
c <- estimateGLMTagwiseDisp(c, design)
fit <- glmFit(c, design)
```

DE for failing atrium groups
```{r}
m<-rep(0,ncol(design))
aa<-grep("failing_atrium_control", colnames(design))
vv<-grep("failing_atrium_repro", colnames(design))
m[aa]<- -1 ### control
m[vv]<- 1 ## experiment
colnames(design)
m
lrt <- glmLRT(fit, contrast = m)
topTags(lrt, n=nrow(lrt$table))->sorted
```

```{r}
head(sorted$table)
```

# End

